---
title: "Confirmatory Factor Analysis"
---

```{r}
#| include: false
source(here::here("_setup.R"))
library(semPlot)
library(tinytable)
```

```{r}
#| include: false
knitr::read_chunk(here::here("notes", "R", "data-exploration.R"))
knitr::read_chunk(here::here("notes", "R", "cfa.R"))
```

```{r}
#| include: false
library(countdown)
library(lavaan)
library(semTools)
```

```{r}
#| label: load-data
#| include: false
```

```{r}
#| label: load-data-sop
#| include: false
```

# From Path Analysis to CFA

## Recap: Path Analysis

- Path analysis models relationships among **observed** (measured) variables
- Assumes variables are measured **without error**

---

But in practice, psychological constructs are rarely measured perfectly:

- Self-oriented perfectionism (SOP) ≠ any single item
- Academic procrastination (PASS) ≠ any single item

. . .

> What if our "X" variables are themselves imperfect measurements of an underlying construct?

## Items of SOP

1 = "strongly disagree", 5 = "strongly agree"

::: {.nonincremental}
- I strive to be as perfect as possible.
- I have a strong need to be perfect.
- I never settle for less than perfection from myself.
- It is important to me to be perfect in everything I attempt.
- I do things perfectly, or I don’t do them at all.
:::

::: aside
From The Big Three Perfectionism Scale (BTPS; Smith et al., 2016)
:::

## Latent Variables {.smaller}

:::: {.columns}

::: {.column width="50%"}

**Path model** (observed)

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3
ly_path <- matrix(c(-1, 1, 0, 0), nrow = 2)
semPaths(lavaanify("NGSE ~ SOP"), what = "path", whatLabels = "hide",
         residuals = FALSE, nCharNodes = 0,
         sizeMan = 14, edge.width = 1.5,
         layout = ly_path)
```

SOP and NGSE are treated as **error-free**

:::

::: {.column width="50%"}

**CFA model** (latent)

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3
cfa_sketch <- "
  SOP =~ SOP1 + SOP2 + SOP3
  NGSE =~ NGSE1 + NGSE2 + NGSE3
  SOP ~~ NGSE
"
semPaths(lavaanify(cfa_sketch), what = "path", whatLabels = "hide",
         residuals = TRUE, nCharNodes = 0,
         sizeMan = 10, sizeLat = 14, edge.width = 1.5)
```

Observed items are **imperfect indicators** of a latent factor

:::

::::

## Usage of CFA

- "Confirm" that a set items reflects on the hypothesized number of factors
- Part of scale validation (e.g., evidence of internal structure, convergent/discriminant validity)
- Suggest directions for scale improvement

# The Common Factor Model

## The Common Factor Model {.smaller}

For item $j$ and factor $F$:

$$
Y_j = \lambda_j \eta + e_j
$$

- $\lambda_j$ = **factor loading**---how strongly item $j$ reflects the factor
- $\eta$ = **latent factor** (unobserved)
- $e_j$ = **unique error**---item-specific variance + measurement error
    * $\text{Var}(e_j) = \theta_j$ is called the **uniqueness**

. . .

::: {.callout-note}

## Communality vs. Uniqueness

$$
\underbrace{\lambda_j^2 \, \text{Var}(\eta)}_{\text{communality}} + \underbrace{\theta_j}_{\text{uniqueness}} = \text{Var}(Y_j)
$$

Communality = proportion of item variance explained by the common factor

:::

## Common Factor Model {.smaller}

### (Matrix form, Multiple Factors)

$$
\begin{bmatrix}
Y_1 \\
Y_2 \\
\vdots \\
Y_p
\end{bmatrix}
=
\begin{bmatrix}
\lambda_{11} & \lambda_{12} & \cdots & \lambda_{1m} \\
\lambda_{21} & \lambda_{22} & \cdots & \lambda_{2m} \\
\vdots & \vdots & \ddots & \vdots \\
\lambda_{p1} & \lambda_{p2} & \cdots & \lambda_{pm}
\end{bmatrix}
\begin{bmatrix}
\eta_1 \\
\eta_2 \\
\vdots \\
\eta_m
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\
e_2 \\
\vdots \\
e_p
\end{bmatrix}
$$

$$
\mathbf{Y} = \mathbf{\Lambda} \boldsymbol{\eta} + \mathbf{e}
$$

Cov($\boldsymbol{\eta}$) = $\boldsymbol{\Psi}_{m \times m}$

Cov($\mathbf{e}$) = $\boldsymbol{\Theta}_{p \times p}$

# The CFA Model

## Implied Covariance Structure

For $p$ indicators of one factor $\eta$:

$$
\boldsymbol{\Sigma} = \boldsymbol{\Lambda} \boldsymbol{\Psi} \boldsymbol{\Lambda}^\top + \boldsymbol{\Theta}
$$

- $\boldsymbol{\Lambda}$ ($p \times 1$): vector of factor loadings
- $\boldsymbol{\Psi}$ ($1 \times 1$): factor variance
- $\boldsymbol{\Theta}$ ($p \times p$, diagonal): uniquenesses

. . .

Just like path analysis, we estimate model parameters from the **covariance matrix** of the observed items.

# Steps of CFA

## Adapted from Bollen (1989; 2026)

1. Model specification
2. Model identification
3. Model estimation
4. Test and evaluation
5. Model modification/respecification

::: {.fragment}
Same steps as path analysis---but step 4 is now more informative because $df > 0$ gives us a global $\chi^2$ test and approximate fit indices.
:::

## 1. Specification

- Decide **which indicators** load on **which factors**
- Decide which cross-loadings or residual covariances to include (or exclude)
- CFA is **confirmatory**: the factor structure comes from theory, not data

. . .

::: {.callout-note}

## CFA vs. EFA

In EFA, all factors load on all items, and rotation is used to find simple structure.

In CFA, the loading pattern is **pre-specified**---absence of an arrow is a *strong* zero constraint.

:::

## 2. Model Identification

> Can we uniquely determine all parameters from the observed covariance matrix?

- Number of unique elements: $p^* = p(p+1)/2$
- Free parameters: 
    * $p$ loadings
    * 1 factor variance
    * $p$ uniquenesses
    * [-1 for scaling constraint]{style="color: red;"}

## Scaling the Latent Variable

Two common strategies for giving the latent variable a unit:

| Strategy | Constraint | Effect |
|---|---|---|
| Marker variable | Fix $\lambda_1 = 1$ | Factor variance $\psi$ is free |
| Standardized LV | Fix $\psi = 1$ | All loadings free |

. . .

::: {.callout-note}
We will use `std.lv = TRUE`---standardized latent variable. All loadings are freely estimated and directly interpretable as regression coefficients onto a standardized factor.
:::

## Degrees of Freedom {visibility="hidden"}

$$
df = p^* - t
$$

- $p^*$ = unique elements in $\boldsymbol{\Sigma}$
- $t$ = number of free parameters
- $df > 0$: **overidentified**---the model makes testable predictions about the data

## Number of Indicators

- With one factor, 3 items give a just-identified model
- Generally speaking, 4+ indicators per factor are recommended (e.g., Kline, 2023)

. . .

::: {.callout-note}
Unlike the just-identified path model ($df = 0$), CFA models are typically overidentified. This enables a **global model fit test**.
:::

## 3. Estimation

- (Approximately) continuous and normal indicators: Maximum Likelihood (ML)
- (Mildly) non-normal indicators or ordinal items with 5 categories: **MLR**
    * Huber–White robust SEs
    * Yuan–Bentler scaled $\chi^2$
- Ordinal indicators with 2-4 categories: Diagonally Weighted Least Squares (DWLS)
- Missing data: **FIML** (`missing = "FIML"`)---uses all available information

# Using the *lavaan* Package

## Model Specification

```{r}
#| label: cfa-1factor
#| echo: true
#| results: false
```

. . .

::: {.callout-note .nonincremental}

## *lavaan* operators for CFA

- `SOP =~ SOP_1` means "SOP **manifests as** SOP_1" (loading $\lambda_1$)
- `SOP_1 ~~ SOP_1` means uniqueness $\theta_1$ (estimated by default)
- `SOP ~~ SOP` is the factor variance, **fixed to 1** by `std.lv = TRUE`

:::

## Estimation Output

```{r}
#| echo: true
summary(fit_1f, fit.measures = TRUE, standardized = TRUE)
```

## Standardized Loadings

```{r}
#| label: loadings-cfa
#| echo: true
#| output-location: column
```

. . .

::: {.callout-note}
Quick rule of thumb: Values $\geq .40$ are generally considered acceptable; $\geq .60$ is good.
:::

## Path Diagram

```{r}
#| echo: true
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
semPaths(fit_1f, whatLabels = "std", layout = "tree",
         edge.label.cex = 1.2, sizeMan = 8, sizeLat = 12,
         nCharNodes = 0, residuals = TRUE)
```

## 4. Model Fit

- Global $\chi^2$ test: can using 1 factor fully explain the covariance among items?
    * $H_0$: Specified model fits the data perfectly, $\boldsymbol{\Sigma} = \boldsymbol{\Sigma}(\hat{\theta})$
- Goodness-of-fit indices: how close is $\boldsymbol{\Sigma}(\hat{\theta})$ to $\boldsymbol{\Sigma}$?
- Local fit: which item pairs are not well explained by the model?
    * E.g., Correlation residuals (see [note on CFA](/notes/cfa.qmd))

## Approximate Fit Indices {.smaller}

```{r}
#| echo: false
#| fig-height: 2
# Set up a clean, blank plotting area
par(mar = c(1, 1, 1, 1), family = "serif")
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 1.5))

# Define X-coordinates for the scale points and Y-coordinate for the baseline
s_x <- 1
sigma_theta_x <- 3
sigma_baseline_x <- 9
y_line <- 0.5

# Draw the linear scale base line
lines(c(s_x, sigma_baseline_x), c(y_line, y_line), lwd = 1)

# Add tick marks
tick_size <- 0.15
segments(s_x, y_line - tick_size, s_x, y_line + tick_size, lwd = 1.5)
segments(sigma_theta_x, y_line - tick_size, sigma_theta_x, y_line + tick_size, lwd = 1.5)
segments(sigma_baseline_x, y_line - tick_size, sigma_baseline_x, y_line + tick_size, lwd = 1.5)

# Add matrix labels below the ticks
label_offset_y <- 0.4
text(s_x, y_line - label_offset_y, expression(bold(Sigma)), font = 2, cex = 1.3)
text(sigma_theta_x, y_line - label_offset_y, expression(bold(Sigma)(theta)), cex = 1.3)
text(sigma_baseline_x, y_line - label_offset_y, expression(bold(Sigma[baseline])), cex = 1.3)

# Draw arrows between the points
arrow_width <- 2
arrow_length <- 0.12

# Red arrow (S to Sigma_theta)
arrows(s_x + 0.01, y_line, sigma_theta_x - 0.01, y_line, 
       col = "red", length = arrow_length, angle = 20, code = 3, lwd = arrow_width)

# Blue arrow (Sigma_theta to Sigma_baseline)
arrows(sigma_theta_x + 0.01, y_line, sigma_baseline_x - 0.01, y_line, 
       col = "#0072B2", length = arrow_length, angle = 20, code = 3, lwd = arrow_width)

# Place RMSEA and SRMR stacked above the red segment
text_y_red1 <- y_line + 0.8
text_y_red2 <- y_line + 0.4
mid_red_x <- mean(c(s_x, sigma_theta_x))

text(mid_red_x, text_y_red1, "RMSEA (< .06)", col = "red", cex = 1.1)
text(mid_red_x, text_y_red2, "SRMR (< .08)", col = "red", cex = 1.1)

# Place CFI above the blue segment
text_y_blue <- y_line + 0.6
mid_blue_x <- mean(c(sigma_theta_x, sigma_baseline_x))

text(mid_blue_x, text_y_blue, "CFI (> .95)", col = "#0072B2", cex = 1.1)
```

```{r}
#| label: fit-indices
#| echo: true
```

. . .

::: {.callout-warning}
These cutoffs (Hu & Bentler, 1999) are **guidelines**, not golden rules. They were derived under specific simulation conditions and do not replace substantive judgment.
:::

## {visibility="hidden"}

Quick-reference guidelines:

| Index | Good fit | Acceptable |
|---|---|---|
| CFI | $\geq .95$ | $\geq .90$ |
| RMSEA | $\leq .06$ | $\leq .10$ |
| SRMR | $\leq .08$ | $\leq .10$ |

## Correlation Residuals {visibility="hidden"}

Difference between observed and model-implied correlations---what the factor model *fails* to explain:

```{r}
#| label: cor-residuals
#| echo: false
```

. . .

::: {.callout-note}
A rule of thumb: residuals $> |0.10|$ may indicate local misfit between a pair of items.
:::

## 5. Model Respecification

Modification Indices (MI): Estimate how much $\chi^2$ would decrease if a fixed parameter were freed:

```{r}
#| label: mod-indices
#| echo: true
```

::: {.callout-warning .fragment}
Only modify the model when there is **theoretical justification**---post-hoc modifications capitalize on chance and should be cross-validated.
:::

## Example Modification {.smaller visibility="hidden"} 

Adding error covariance between SOP_1 and SOP_2:

```{r}
#| echo: true
model_1f_mod <- "
  SOP =~ SOP_1 + SOP_2 + SOP_3 + SOP_4 + SOP_5
  SOP_1 ~~ SOP_2
"
```

```{r}
#| echo: false
fit_1f_mod <- cfa(model_1f_mod, data = sop_data,
                  std.lv = TRUE, estimator = "MLR")
semPaths(fit_1f_mod, whatLabels = "std", layout = "tree",
         edge.label.cex = 1.2, sizeMan = 8, sizeLat = 12,
         nCharNodes = 0, residuals = TRUE)
```


##

::: {.callout-tip}

## Reporting Sample

Self-oriented perfectionism was assessed with five items from the SOP scale (Williams & Edwards, 2021), rated on a 5-point Likert scale. We fit a unidimensional confirmatory factor analysis (CFA) model using the `lavaan` package (v`r packageVersion("lavaan")`; Rosseel, 2012) in R. The latent factor variance was fixed to 1 for identification (`std.lv = TRUE`). To account for mild non-normality in the item responses, we used robust maximum likelihood estimation (MLR) with Yuan–Bentler scaled standard errors and a Satorra–Bentler scaled $\chi^2$ test statistic. Missing data were handled with full information maximum likelihood (FIML). Model fit was evaluated using the robust CFI ($\geq .95$), robust RMSEA ($\leq .06$), and SRMR ($\leq .08$; Hu & Bentler, 1999). 

The hypothesized unidimensional CFA model showed acceptable fit to the data, robust $\chi^2$(`r fitMeasures(fit_1f, "df")`) = `r round(fitMeasures(fit_1f, "chisq.scaled"), 2)`, $p$ `r ifelse(fitMeasures(fit_1f, "pvalue.scaled") < .001, "< .001", paste0("= ", round(fitMeasures(fit_1f, "pvalue.scaled"), 3)))`, robust CFI = `r round(fitMeasures(fit_1f, "cfi.robust"), 3)`, robust RMSEA = `r round(fitMeasures(fit_1f, "rmsea.robust"), 3)`, 90% CI [`r round(fitMeasures(fit_1f, "rmsea.ci.lower.robust"), 3)`, `r round(fitMeasures(fit_1f, "rmsea.ci.upper.robust"), 3)`], SRMR = `r round(fitMeasures(fit_1f, "srmr"), 3)`. Inspection of modification indices did not reveal any theoretically justified model modifications. All five items loaded significantly on the SOP factor, with standardized loadings ranging from `r round(min(standardizedSolution(fit_1f) |> dplyr::filter(op == "=~") |> dplyr::pull(est.std)), 2)` to `r round(max(standardizedSolution(fit_1f) |> dplyr::filter(op == "=~") |> dplyr::pull(est.std)), 2)`. 

:::

## 5. Composite Reliability {visibility="hidden"}

Proportion of composite score variance that is **true score variance**:

```{r}
#| label: composite-reliability
#| echo: true
```

. . .

::: {.callout-note}

## McDonald's $\omega$

$$
\omega = \frac{\left(\sum_{j=1}^{p} \lambda_j\right)^2 \text{Var}(F)}{\left(\sum_{j=1}^{p} \lambda_j\right)^2 \text{Var}(F) + \sum_{j=1}^{p} \theta_j}
$$

This is the reliability of the **observed composite** (e.g., sum score), not of the latent variable itself.

:::

## Exercise {background-color='{{< brand color fed-teal >}}' .smaller}

::: {.nonincremental}
1. Run the R code in "note/data-exploration.R", and then the code in "note/cfa.R" to fit a CFA model.
    a. Which item has the highest standardized loading?
2. Fit the following modified CFA model:

   ```r
   model_1f_mod <- "
     SOP =~ SOP_1 + SOP_2 + SOP_3 + SOP_4 + SOP_5
     SOP_1 ~~ SOP_2
   "
   fit_1f_mod <- cfa(model_1f_mod, data = sop_data,
                     std.lv = TRUE, estimator = "MLR")
   ```

    a. How much did the fit improve in RMSEA and CFI?
    b. What was the standardized covariance between SOP_1 and SOP_2?
:::

`r countdown(minutes = 15)`
