---
title: "Full SEM"
---

```{r}
#| include: false
source(here::here("_setup.R"))
library(semPlot)
library(semptools)
library(ggdag)
library(tinytable)
library(modelsummary)
library(lavaan)
library(semTools)
library(countdown)
library(dplyr)
```

```{r}
#| include: false
knitr::read_chunk(here::here("notes", "R", "data-exploration.R"))
knitr::read_chunk(here::here("notes", "R", "structural.R"))
```

```{r}
#| label: load-data
#| include: false
```

```{r}
#| label: we-parcels
#| include: false
```

# From CFA to Full SEM

## What is Full SEM?

SEM = CFA (measurement) + Path analysis (structural)

- **Measurement model**: how latent variables are measured by indicators
- **Structural model**: directional relations among latent variables

. . .

Unlike observed-score path models, SEM explicitly accounts for measurement error.

## An Example of Full SEM

```{r}
#| label: we-sr
#| include: false
```

```{r}
p <- semPaths(
    lavaanify(model_sr_we),
    style = "ram",
    layout = "tree2",
    rotation = 2,
    nCharNodes = 0,
    sizeMan = 5,
    sizeLat = 9,
    edge.label.cex = 0.7,
    curvePivot = TRUE,
    fade = FALSE,
    DoNotPlot = TRUE
)

factor_layout <- matrix(
    c(NA, "NGSE", NA,
      "SOP", NA, "PASS"),
    nrow = 2, byrow = TRUE
)

indicator_push <- c(
    SOP = 1.5,
    NGSE = 1.5,
    PASS = 1.5
)
loading_position <- c(NGSE = 0.7)

p2 <- set_sem_layout(
    p,
    indicator_push  = indicator_push,
    loading_position = loading_position,
    factor_layout    = factor_layout,
    factor_point_to  = c(SOP = "left", NGSE = "up", PASS = "right"),
    indicator_spread = c(SOP = 1.4, NGSE = 2.5, PASS = 1.2)
)

plot(p2)
```

## Why Use Full SEM?

- We can test mediation or direct effects among **latent** constructs
- Path coefficients are less attenuated by measurement error
- Model fit evaluates whether the full covariance structure is plausible

. . .

::: {.callout-note}
SEM is especially useful when your substantive hypotheses are causal but your key variables are latent.
:::

# Example: SOP, NGSE, and PASS

## Example Data {.smaller}

- **Data source**: Williams & Edwards ([2021](https://cipp.ug.edu.pl/Conceptual-replication-of-Seo-2008-Self-efficacy-as-a-mediator-in-the-relationship,142208,0,2.html), conceptual replication of Seo, 2008)
- **Sample size**: $N = 575$
- Constructs:
    - `SOP`: Self-Oriented Perfectionism (5 items)
    - `NGSE`: General Self-Efficacy (8 items)
    - `PASS`: Academic Procrastination (4 parcels: assignments, exams, reading, general)

## Hypothesized Mediation Model

```{r}
#| label: fig-dag
#| echo: false
#| fig-width: 5
#| fig-height: 3
```

$$
\begin{aligned}
\text{NGSE} &= a\cdot\text{SOP} + \zeta_1 \\
\text{PASS} &= b\cdot\text{NGSE} + c'\cdot\text{SOP} + \zeta_2
\end{aligned}
$$

## Two-Step SEM Workflow

1. **Step 1: Full Measurement model (CFA)**
    - Verify indicators measure intended factors
    - Check global fit and loading pattern
2. **Step 2: Structural model (SR/SEM)**
    - Impose directional paths among latent factors
    - Evaluate direct, indirect, and total effects

. . .

::: {.callout-note}
If Step 1 is poor, Step 2 interpretations are not trustworthy.
:::

# Step 1: Measurement Model

## Data Overview {.smaller .scrollable}

```{r}
#| echo: false
f <- function(d, ...) {
    function() {
        hist(
            d,
            axes = FALSE,
            ann = FALSE,
            col = "#005f96",
            breaks = if (length(unique(d)) <= 7) 0:max(d) else "Sturges"
        )
    }
}
modelsummary::datasummary_skim(
    we_data,
    fun_numeric = list(
        Mean = Mean,
        SD = SD,
        Min = Min,
        Median = Median,
        Max = Max,
        Histogram = \(x) ""
    )
) |>
    tinytable::plot_tt(j = "Histogram", fun = f, data = we_data) |>
    tinytable::style_tt(fontsize = 0.8)
```

## Correlation Structure {.smaller}

```{r}
#| label: we-cor-heatmap
#| echo: false
```

## CFA Model Specification

```{r}
#| label: we-cfa
#| echo: true
#| results: false
```

## {.smaller}

### CFA Fit

```{r}
#| echo: false
fm_cfa <- fitMeasures(fit_cfa_we, c("chisq.scaled", "df", "cfi.robust", "rmsea.robust", "srmr"))
fit_cfa_tbl <- tibble::tibble(
  Model = "CFA",
  `Chisq` = fm_cfa["chisq.scaled"],
  `df` = fm_cfa["df"],
  `CFI*` = fm_cfa["cfi.robust"],
  `RMSEA*` = fm_cfa["rmsea.robust"],
  SRMR = fm_cfa["srmr"]
)

fit_cfa_tbl |>
  mutate(across(-Model, round, 3)) |>
  tt()
```

### Standardized Loadings 

```{r}
#| echo: false
#| layout-ncol: 3
tbl_loadings <- function(lhs, fit = fit_cfa_we) {
  standardizedSolution(fit) |>
    filter(op == "=~", lhs == !!lhs) |>
    select(ind = rhs, loading = est.std) |>
    mutate(loading = round(loading, 3)) |>
    tt()
}
tbl_loadings("SOP")
tbl_loadings("NGSE")
tbl_loadings("PASS")
```

# Step 2: Structural Model

## SR Model Specification

```{r}
#| label: we-sr
#| echo: true
#| results: false
#| eval: false
```

## Structural Model Diagram

```{r}
#| label: fig-sr-sempaths
#| echo: false
#| fig-width: 8.5
#| fig-height: 5.3
```

## Structural Paths and Indirect Effect {.smaller}

```{r}
#| echo: false
parameterEstimates(fit_sr_we, standardized = TRUE) |>
  filter(op %in% c("~", ":=")) |>
  select(lhs, op, rhs, est, se, z, pvalue, std.all) |>
  mutate(across(c(est, se, z, pvalue, std.all), round, 3)) |>
  tt()
```

## Fit Comparison: Step 1 vs Step 2

```{r}
#| echo: false
fm_sr <- fitMeasures(fit_sr_we, c("chisq.scaled", "df", "cfi.robust", "rmsea.robust", "srmr"))
fit_compare <- tibble::tibble(
  Model = c("Step 1: CFA", "Step 2: SEM"),
  `Chisq*` = c(fm_cfa["chisq.scaled"], fm_sr["chisq.scaled"]),
  `df` = c(fm_cfa["df"], fm_sr["df"]),
  `CFI*` = c(fm_cfa["cfi.robust"], fm_sr["cfi.robust"]),
  `RMSEA*` = c(fm_cfa["rmsea.robust"], fm_sr["rmsea.robust"]),
  SRMR = c(fm_cfa["srmr"], fm_sr["srmr"])
)

fit_compare |>
  mutate(across(-Model, round, 3)) |>
  tt()
```

## Interpretation of Effects

- **Indirect effect**: $a \times b$ (SOP $\rightarrow$ NGSE $\rightarrow$ PASS)
- **Direct effect**: $c'$ (SOP $\rightarrow$ PASS, controlling for NGSE)
- **Total effect**: $c' + a\times b$

. . .

::: {.callout-note}
If the 95% CI for the indirect effect excludes 0, the mediation effect is statistically supported.
:::

## Brief Causal Caution

- A fitted SEM does **not** prove causal direction
- Causal interpretation depends on assumptions (no key omitted confounders)
- Equivalent/reversed models may show similar global fit

. . .

::: {.callout-warning}
Use theory and design (not fit indices alone) to justify directional claims.
:::

## Summary

- Structural regression combines CFA and path analysis in one model
- Use the two-step strategy: validate measurement first, then test structure
- Indirect effects are straightforward in *lavaan* using `:=`
- Good fit supports plausibility, but does not establish causality by itself

## Exercise {background-color='{{< brand color fed-teal >}}' .smaller}

::: {.nonincremental}
1. Fit the measurement model using `we_data` and check fit.
    a. Which factor has the strongest average standardized loading?
2. Fit the structural mediation model without the direct path (c'):

   ```r
   model2_sr_we <- "
     SOP  =~ SOP_1 + SOP_2 + SOP_3 + SOP_4 + SOP_5
     NGSE =~ NGSE_1 + NGSE_2 + NGSE_3 + NGSE_4 +
             NGSE_5 + NGSE_6 + NGSE_7 + NGSE_8
     PASS =~ Pasgn + Pexam + Pread + Pgen

     NGSE ~ a * SOP
     PASS ~ b * NGSE

     indirect := a * b
   "
   fit2_sr_we <- sem(model2_sr_we, data = we_data,
                     std.lv = TRUE, estimator = "MLR", missing = "FIML")
   ```

    a. Compare fit to the model with the direct path:

    ```r
    anova(fit_sr_we, fit2_sr_we)
    ```
:::

`r countdown(minutes = 15)`

##

<iframe src="https://app.sli.do/event/4yTrQWkJQSZ9CqXeecr8ap" height="100%" width="100%" frameBorder="0" style="min-height: 560px;" allow="clipboard-write" title="Slido"></iframe>