---
title: "Confirmatory Factor Analysis With Categorical Indicators"
---

```{r}
#| label: setup
#| message: false
library(here)
library(lavaan)
library(dplyr)
library(modelsummary)
```

The data come from this paper: <https://doi.org/10.1111/desc.12918>. The authors kindly shared the data on OSF: <https://osf.io/hq4xy/overview>.

```{r}
# Download data from OSF
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}
# Raw items
tiego_raw_name <- here("data", "tiego_raw.csv")
if (!file.exists(tiego_raw_name)) {
    download.file(
        "https://osf.io/download/mtnqj/",
        destfile = tiego_raw_name
    )
}
```

### Load Data

```{r}
#| label: load-tiego-data
#| output: false
tiego_raw <- read.csv(tiego_raw_name)
# Notice that the minimum is -999 for many variables, so -999 is likely a missing code
# Treat -999 as NA
tiego_raw <- read.csv(tiego_raw_name, na.strings = -999)
# Looks like the CBCL items also have some other missing codes
tiego_raw |>
    select(starts_with("CBCLProb")) |>
    glimpse()
# Recode 9 to NA as well
tiego_raw <- tiego_raw |>
    mutate(across(starts_with("CBCLProb"),
                  ~ ifelse(.x == 9, NA, .x)))
```

The data contain items from CBCL (Child Behavior Checklist). The items are rated on a 3-point scale (0 = not true, 1 = somewhat or sometimes true, 2 = very true or often true), and should not be treated as continuous. We will focus on the 7-item DSM-oriented Attention Deficit/Hyperactivity Problems scale, with the seven items being Items 4, 8, 10, 41, 78, 93, 104.

```{r}
# Gather the items into a smaller data set
tiego_adhd <- tiego_raw |>
    select(
        adhd1 = CBCLProb4,  # fails to finish
        adhd2 = CBCLProb8,  # concentrate
        adhd3 = CBCLProb10,  # sit still
        adhd4 = CBCLProb41,  # impulsive
        adhd5 = CBCLProb78,  # inattentive
        adhd6 = CBCLProb93,  # talk much
        adhd7 = CBCLProb104  # loud
    ) |>
    # recode to be ordinal
    mutate(
        across(
            everything(),
            ~ ordered(.x, levels = c(0, 1, 2),
                      labels = c("not true", "somewhat or sometimes true",
                                 "very true or often true")),
            .names = "{.col}_c"))
datasummary_skim(tiego_adhd[, 8:14])  # check the recoded items
```

## Background on Categorical Data

Many scales in psychology, like the CBCL, use Likert-type response formats (e.g., 0-1-2). Standard CFA using Maximum Likelihood (ML) assumes that the observed variables are:

1.  **Continuous**: Can take any value within a range.
2.  **Normally Distributed**: Follow a bell curve.

When we have ordinal data with few categories (e.g., fewer than 5), treating them as continuous can lead to:

-   Biased parameter estimates (e.g., underestimated factor loadings).
-   Incorrect standard errors and $\chi^2$ statistics.
-   Spurious multidimensionality (factors defined by item difficulty rather than content).

### Underlying Latent Response Formulation

Instead of analyzing the discrete scores directly (0, 1, 2), we assume that for each observed categorical item $Y$, there is an underlying, unobserved continuous variable $Y^*$ (Y-star). The observed score is determined by where $Y^*$ falls relative to certain **thresholds** ($\tau$):

$$
Y = \begin{cases} 
0 & \text{if } Y^* \le \tau_1 \\
1 & \text{if } \tau_1 < Y^* \le \tau_2 \\
2 & \text{if } Y^* > \tau_2
\end{cases}
$$

```{r}
#| echo: false
#| fig-height: 3.5
#| code-fold: true
library(ggplot2)
library(patchwork)

tau <- c(-0.2, 1.0)
xpos_text <- (c(-2, tau) + c(tau, 2)) / 2

# Plot 1: Latent density
p1 <- ggplot(data.frame(x = c(-3, 3)), aes(x = x)) +
    stat_function(fun = dnorm) +
    geom_vline(xintercept = tau, linetype = "dashed") +
    annotate("text", x = tau + 0.05, y = 0.3, label = c(expression(tau[1]), expression(tau[2])), hjust = 0) +
    annotate("text", x = xpos_text, y = 0.1, label = 
    c("Y = 0", "Y = 1", "Y = 2")) +
    labs(x = expression(Y^"*"), y = "Density") +
    theme_classic()

# Plot 2: Observed probabilities
probs <- c(pnorm(tau), 1) - c(0, pnorm(tau))
df_bar <- data.frame(
    Category = factor(c("0", "1", "2")),
    Probability = probs
)

p2 <- ggplot(df_bar, aes(x = Category, y = Probability)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = round(Probability, 2)), vjust = -0.5) +
    labs(x = "Observed Category (Y)", y = "Probability") +
    theme_classic()

# Combine plots horizontally
p1 + p2
```

### Polychoric Correlations

Because we care about the relationships between the underlying latent responses ($Y^*$), we estimate the correlations between these $Y^*$ variables rather than the Pearson correlations of the observed integers. These correlations are called **polychoric correlations**. They are typically stronger than the Pearson correlations calculated on coarse categorical data.

### WLSMV Estimator

For categorical CFA, we typically use the **WLSMV** (Weighted Least Squares Mean and Variance adjusted) estimator. 

-   **DWLS**: The core estimation method is Diagonally Weighted Least Squares (DWLS). Unlike Full WLS which requires the inversion of a massive weight matrix (dimensions increase rapidly with more items), DWLS only uses the diagonal elements of the weight matrix. This makes it much more computationally stable and feasible for larger models and smaller sample sizes.
-   **MV**: It then applies robust corrections to the standard errors and test statistics (like $\chi^2$) to account for the fact that the full weight matrix was not used. Specifically, it adjusts both the Mean and Variance of the test statistic distribution.

## (Not Recommended) Treating Items as Continuous

```{r}
fit_cont <- cfa(
    " ADHD =~ adhd1 + adhd2 + adhd3 + adhd4 + adhd5 + adhd6 + adhd7 ",
    data = tiego_adhd, missing = "fiml", estimator = "MLR"
)
summary(fit_cont, fit.measures = TRUE, standardized = TRUE)
```

## Ordinal CFA

The ordinal CFA approach is based on the polychoric correlations among the items:

```{r}
psych::polychoric(tiego_adhd[, 1:7])
```

```{r}
fit_ord <- cfa(
    " ADHD =~ adhd1_c + adhd2_c + adhd3_c + adhd4_c + adhd5_c + adhd6_c + adhd7_c ",
    data = tiego_adhd, estimator = "WLSMV", ordered = TRUE
)
summary(fit_ord, fit.measures = TRUE, standardized = TRUE)
```

The fit indices are different from those in the continuous CFA, and does not appear adequate. We can check the residuals and modification indices to see where the model misfits.

```{r}
resid(fit_ord, type = "cor.bentler")$cov
modindices(fit_ord, sort. = TRUE, minimum.value = 10)
```

### Updated Models

1. Add unique covariance between `adhd1_c` and `adhd2_c` and between adhd6_c and adhd7_c based on modification indices.

::: {.callout-caution}

## Reminder

When adding unique covariances, make sure they make theoretical sense. Also, avoid adding too many unique covariances post hoc.

:::

```{r}
fit_ord2 <- cfa(
    " ADHD =~ adhd1_c + adhd2_c + adhd3_c + adhd4_c + adhd5_c + adhd6_c + adhd7_c 
      adhd1_c ~~ adhd2_c
      adhd6_c ~~ adhd7_c ",
    data = tiego_adhd, estimator = "WLSMV", ordered = TRUE
)
summary(fit_ord2, fit.measures = TRUE, standardized = TRUE)
```

