---
title: "(Bnous) Another Example of Confirmatory Factor Analysis"
---

```{r}
#| label: setup
#| message: false
library(lavaan)
library(semPlot)
library(semptools)
library(lavaangui)
library(dplyr)
library(semTools)
library(modelsummary)
library(tinytable)
```

```{r}
#| include: false
knitr::read_chunk("R/data-exploration-asags.R")
knitr::read_chunk("R/cfa-asags-3factor.R")
knitr::read_chunk("R/cfa-asags-bifactor.R")
```

```{r}
#| label: load-data
#| message: false
```

## Dataset: Attainment of School Achievement Goal Scale (A-SAGS)

The **A-SAGS** (Boivin et al., 2026) measures the perceived attainment of three types of achievement goals in academic settings. Participants rated 13 items on a **7-point Likert scale** (1 = *not at all* to 7 = *completely*).

| Factor | Items | Description |
|--------|-------|-------------|
| **MGA** (Mastery Goal Attainment) | ga1, ga4, ga7, ga10, ga13 | Learning and developing competence |
| **SRGA** (Social Reference Goal Attainment) | ga2, ga5, ga8, ga11 | Outperforming others |
| **NGA** (Nurturance Goal Attainment) | ga3, ga6, ga9, ga12 | Helping others succeed |

The three subscales are hypothesized to reflect a single higher-order construct: **Goal Attainment (GA)**. The dataset has *N* = 527 with some missing values.

### Descriptive Statistics

```{r}
#| label: descriptives
```

### Inter-item Correlations

```{r}
#| label: cor-heatmap
```

::: {.callout-note}
Blue cells indicate positive correlations, red cells negative, intensity reflects magnitude. Items within the same subscale (MGA: ga1, ga4, ga7, ga10, ga13; SRGA: ga2, ga5, ga8, ga11; NGA: ga3, ga6, ga9, ga12) should cluster together.
:::

### Dimensionality

Before imposing a factor structure via CFA, it is useful to get a rough sense of how many dimensions underlie the 13 items. We inspect **eigenvalues** of the correlation matrix and run a quick **parallel analysis**.

```{r}
#| label: fig-scree
#| fig-cap: "Scree plot with parallel analysis reference line. Factors above the red line are retained."
```

```{r}
#| label: eigenvalues
```

::: {.callout-note}
Parallel analysis retains factors whose eigenvalues exceed those expected from random data. The eigenvalue-greater-than-1 rule (Kaiser criterion) tends to over-extract and should be treated as an upper bound only.
:::

The A-SAGS was designed with **3 first-order factors** plus an overarching **goal attainment** dimension, so we expect 1–3 meaningful factors. We proceed to CFA to evaluate this structure formally.

# Confirmatory Factor Analysis (Chapter 13)

Confirmatory Factor Analysis (CFA) is a measurement model where we specify how observed variables (indicators) relate to underlying latent constructs (factors). Unlike EFA, CFA requires a priori specification of the factor structure.

## Key Concepts

- **Latent Variables (Factors):** Unobserved constructs (e.g., goal attainment) that cause observed scores.
- **Indicators:** Observed variables (e.g., survey items) used to measure the latent construct.
- **Factor Loadings ($\lambda$):** Regression coefficients representing the strength of the relationship between the factor and the indicator.
- **Measurement Errors ($\epsilon$):** The portion of the indicator variance not explained by the factor.

## Identification in CFA

To estimate a CFA model, we must set the scale for the latent variable. Two common methods:

1.  **Marker Variable:** Fix one factor loading to 1.0 (default in `lavaan`).
2.  **Standardized Latent Variable:** Fix the factor variance to 1.0 (used here with `std.lv = TRUE`).

## A Note on Estimation

The A-SAGS items are technically ordinal, but with 7 response categories they can reasonably be treated as approximately continuous. We use **MLR** (Maximum Likelihood with Robust standard errors), which:

- Treats items as continuous (appropriate for ≥ 5 ordered categories)
- Provides Satorra–Bentler scaled $\chi^2$ and robust standard errors that are robust to non-normality
- Produces AIC and BIC for model comparison

::: {.callout-note}
For fewer response categories (≤ 4), or when the distribution is highly skewed, prefer `ordered = TRUE` with `estimator = "WLSMV"` instead.
:::

## Example 1: First-Order CFA (3 Factors)

We first fit a correlated three-factor model to evaluate the measurement structure of each subscale before introducing the higher-order factor.

```{r}
#| label: cfa-3factor
```

```{r}
#| label: cfa-3factor-plot
#| output: false
```

```{r}
#| label: cfa-3factor-plot-styled
```

## Example 2: Bi-factor CFA

The **bi-factor model** is an alternative to the hierarchical model for representing a general factor alongside specific factors. Instead of routing the general factor *through* the subscales, the bi-factor model has:

- A **general factor (GA)** loading directly on all 13 items
- Three **specific factors** (MGA, SRGA, NGA) each loading on their respective items
- All factors constrained to be **orthogonal** (uncorrelated)

This allows the model to partition item variance into general and specific components simultaneously, rather than assuming the general factor operates solely through the subscale factors.

In `lavaan`, bi-factor models are specified with `bifactor()` inside `cfa()`:

```{r}
#| label: cfa-bifactor
```
```{r}
#| label: cfa-bifactor-plot
#| eval: false
```
::: {.callout-note}

## Hierarchical vs. Bi-factor

| | Hierarchical | Bi-factor |
|---|---|---|
| General factor | Acts *through* subscales | Acts *directly* on items |
| Specific factors | Residual variance of subscales | Residual variance of items after removing general factor |
| Factor correlations | General ↔ specific: fixed (via structure) | All factors orthogonal |
| df | More constrained | Same or fewer constraints |

The bi-factor model is more appropriate when you believe the general factor directly influences item responses, not just indirectly through group factors.

:::

## Model Fit

### Correlation Residuals

The factor model assumes that the latent factors fully explain the correlations among the indicators. Residuals > |0.10| suggest local misfit.

```{r}
#| label: cor-residuals
```

### Modification Indices

Modification indices (MIs) estimate how much the overall $\chi^2$ would decrease if a currently fixed parameter were freed.

```{r}
#| label: mod-indices
```

### Goodness of Fit Indices

#### Exact Fit

- **Model Chi-Square ($\chi^2$)**
    * Tests the null hypothesis that the model perfectly reproduces the population covariance matrix.
    * **Limitation**: Highly sensitive to sample size; rejection does not indicate the *degree* of misfit.

#### Approximate Fit Indices

- **RMSEA**: Misfit per degree of freedom, adjusted for model complexity.  
  Cutoffs: < 0.05 (close), 0.05–0.08 (reasonable), > 0.10 (poor).  
  Includes 90% CI and test of close fit (H₀: RMSEA ≤ 0.05).

- **CFI**: Compares the target model to the independence (null) baseline model.  
  ≥ 0.95 indicates good fit; less affected by sample size than $\chi^2$.

- **SRMR**: Average standardized residual between observed and model-implied correlations.  
  < 0.08 indicates acceptable fit.

::: {.callout-note}

**Robust Fit Indices with MLR**

With `estimator = "MLR"`, lavaan computes several versions of each index:

| Suffix | Based on | Recommendation |
|--------|----------|----------------|
| (none) | Naive ML $\chi^2$ | Avoid; ignores non-normality |
| `.scaled` | Satorra–Bentler scaled $\chi^2$ | Common but suboptimal |
| `.robust` | Yuan–Bentler $T_2^*$ | **Preferred** |

Savalei & Rosseel recommend the `.robust` variants because the Yuan–Bentler $T_2^*$ statistic yields better-behaved fit indices under non-normality than the Satorra–Bentler correction. In lavaan these are `cfi.robust`, `tli.robust`, `rmsea.robust`, etc.

:::

::: {.callout-warning}

**Cutoff Values Should Not Be Used Mechanically**

- Cutoffs are rough guidelines, not absolute rules.
- Consider sample size, model complexity, and estimation method.
- A model with "good" fit indices can be theoretically meaningless.
- Always examine residuals, modification indices, and parameter estimates.

:::

#### Comparing Models

The first-order and bi-factor models are **not nested**, so we compare them using information criteria (AIC/BIC) rather than a $\chi^2$ difference test:

```{r}
#| label: lrt-compare
```

#### Extracting Fit Indices

```{r}
#| label: fit-indices
```

### Quick Table of Results

```{r}
#| label: fit-summary-table
```

## (Predicted) Factor Scores and Diagnostics

```{r}
#| label: factor-scores
```

### Residuals: Indicator − Predicted Indicator Scores

```{r}
#| label: fig-bi-residuals
#| fig-cap: "Indicator residuals from the bi-factor CFA model."
#| fig-width: 9
#| fig-height: 7
```

## Composite Reliability

```{r}
#| label: composite-reliability
```

## References

Boivin, T., Veilleux, C., Egerton-Graham, M., Caron, P.-O., & Gaudreau, P. (2026). *A tutorial on evaluating confirmatory factor analysis with lavaan and dynamic: Integrating statistical and conceptual approaches*. PsyArXiv. <https://doi.org/10.31234/osf.io/2m6rk_v4>

Kline, R. B. (2023). *Principles and practice of structural equation modeling* (5th ed.). Guilford Press.
