---
title: "Structural Models"
---

```{r}
#| label: setup
#| message: false
library(lavaan)
library(lavaangui)
library(semPlot)
library(semptools)
library(semTools)
library(ggdag)
library(ggplot2)
library(dplyr)
library(modelsummary)
library(tinytable)
source(here::here("notes", "R", "sem_table.R"))
```

```{r}
#| include: false
knitr::read_chunk("R/data-exploration.R")
knitr::read_chunk("R/structural.R")
```

# Introduction (Chapter 15)

Structural Regression (SR) models, also known as **Structural Equation Modeling (SEM)** in the broad sense (or "Latent Variable Path Analysis"), combine two components:

1.  **Measurement Model**: Confirmatory Factor Analysis (CFA) describing how latent variables are measured by observed indicators.
2.  **Structural Model**: Path analysis describing the causal relationships (regressions) among the latent variables.

$$
\text{SEM} = \text{CFA} + \text{Path Analysis}
$$

## Causal Assumptions in SR Models

Every directed arrow in the structural model represents a **hypothesized causal effect**, not merely a statistical association. By drawing an arrow from $X$ to $Y$, the researcher claims that $X$ influences $Y$---not only that they tend to move together. The structural model is therefore best understood as a **Directed Acyclic Graph (DAG)** extended to include latent variables, and the same principles of causal identification apply. In structural models or causal models, the main challenge is to identify whether variables **not in the model**, either measured or unmeasured, could bias the estimated path coefficients. There are three main types of variables:

**Confounder.** An unmeasured common cause $U$ of $X$ and $Y$ creates a spurious back-door path:
$$
X \leftarrow U \rightarrow Y
$$
The estimated path coefficient then conflates the direct effect of $X$ with the spurious association transmitted through $U$. Blocking this path requires including appropriate observed covariates or using an instrumental variable.

**Mediator.** Including a variable that lies *on* the hypothesized causal pathway ($X \to M \to Y$) as a covariate when estimating $X \to Y$ blocks part of $X$'s genuine effect. This is why regression assumptions (no causal order among predictors) cannot be applied uncritically when variables are causally connected.

**Collider.** Conditioning on a variable caused by two others (a *collider*, e.g., $X \to C \leftarrow Y$) opens a spurious association between those two causes. This can arise when selecting participants on a common outcome or when a descendant of a collider is included as a covariate.

The key discipline from causal inference:

> **Identify before you estimate.** Causal assumptions must be justified with subject-matter knowledge *before* fitting the model. The data alone cannot distinguish a correctly specified causal model from a misspecified one.

When causal structure is uncertain, tools such as `dagitty` can enumerate minimal sufficient adjustment sets and derive the testable conditional independencies implied by the DAG---providing an empirical check on at least some aspects of the assumed structure.

## The Two-Step Approach

Many SEM scholars (e.g., Kline, 2023) strongly recommend a two-step modeling strategy:

1.  **Respecify and test the measurement model**: specificy a CFA model where all latent variables are allowed to correlate (i.e., saturated structural part).
    *   Ensure factors are well-measured.
    *   Check model fit.
    *   If the measurement model is bad, the SR model will effectively necessarily be bad or misleading.
2.  **Test the structural model**: Impose the hypothesized structural paths (regressions) among the latent variables.

# Example: Self-Oriented Perfectionism, Self-Efficacy, and Procrastination

This example uses data from Williams & Edwards (2021) to demonstrate a **mediation structural regression model**. The hypothesized model follows Seo (2008):

```{r}
#| label: fig-dag
#| fig-cap: "Hypothesized mediation model with path labels."
#| fig-width: 5
#| fig-height: 3
#| message: false
```

$$
\begin{aligned}
\text{NGSE}      &= a \cdot \text{SOP} + \zeta_1 \\
\text{PASS} &= b \cdot \text{NGSE} + c' \cdot \text{SOP} + \zeta_2
\end{aligned}
$$

where SOP is self-oriented perfectionism, NGSE is general self-efficacy (Chen et al., 2001), and PASS is academic procrastination.

## Data and PASS Parcels

The data are downloaded from OSF using the same exclusion pipeline as the CFA note.

```{r}
#| label: load-data
#| message: false
#| include: false
```

Rather than using all 12 PASS items as indicators of a single factor, Williams & Edwards formed **item parcels**---one sum score per content area:

```{r}
#| label: we-parcels
#| include: false
```

::: {.callout-note}
## Item Parceling

**Parceling** aggregates multiple items into composite scores before fitting the SEM. Benefits include fewer parameters, a better indicator-to-sample-size ratio, and composites that are more normally distributed than raw ordinal items. The four parcels (assignments, exams, reading, general) replace 12 item indicators with 4 content-balanced composites for the PASS factor.
:::

### Descriptive Statistics

```{r}
#| label: we-descriptives
```

### Inter-item Correlations

```{r}
#| label: we-cor-heatmap
```

::: {.callout-note}
Blue cells = positive correlations, red = negative; intensity = magnitude. SOP items should correlate positively with each other, NGSE items with each other, and the PASS parcels with each other, while cross-scale correlations reflect the hypothesized structural relationships.
:::

## Step 1: Measurement Model

We first fit all three factors simultaneously with a saturated structural part (all factors freely correlated) to check that the measurement model is acceptable before imposing the mediation constraints.

```{r}
#| label: we-cfa
```

## Step 2: Structural Regression Model

We constrain the structural part to the hypothesized mediation model. Parameter labels (`a`, `b`, `c`) allow us to define the indirect and total effects using the `:=` operator.

```{r}
#| label: we-sr
```

## Indirect Effect

The indirect effect (SOP → NGSE → PASS) with its 95% CI from the Monte Carlo method:

```{r}
#| label: we-indirect
```

::: {.callout-note}
The **Monte Carlo method** (MacKinnon et al., 2004) draws repeated samples from the joint sampling distribution of $a$ and $b$ (using their estimated means and covariance matrix) and computes $a \times b$ for each draw, yielding an empirical CI. It handles the asymmetric distribution of a product better than the delta method while requiring no re-estimation of the model, unlike bootstrapping.
:::

## Causal Assumptions for the Mediation Interpretation

Reporting $a \times b$ as a *causal* indirect effect requires that the structural model satisfies a set of **no-unmeasured-confounding** conditions. Three assumptions must hold simultaneously:

1. **No unmeasured confounding of SOP → NGSE**: No latent variable simultaneously causes both self-oriented perfectionism and general self-efficacy.
2. **No unmeasured confounding of SOP → PASS**: No latent variable causes both self-oriented perfectionism and academic procrastination through a path not captured by the model.
3. **No unmeasured confounding of NGSE → PASS** (*Sequential Ignorability*): No latent variable $U$ simultaneously causes self-efficacy and academic procrastination. This is the most critical assumption---and the hardest to defend. Personality traits, cognitive habits, or prior academic experiences could plausibly affect both NGSE and PASS independently of SOP.

```{r}
#| label: fig-mediation-assumptions-dag
#| fig-cap: "Mediation DAG with a latent confounder U between NGSE and PASS. Sequential ignorability requires that no such U exists."
#| message: false
#| fig-width: 5
#| fig-height: 3
```

Because SOP is measured rather than experimentally manipulated, none of these three conditions is guaranteed by design. The indirect effect should therefore be interpreted as a *conditional association consistent with the mediation hypothesis*, not as a confirmed causal mechanism. A sensitivity analysis---asking how large an unmodeled correlation between the disturbances of NGSE and PASS would need to be to nullify the indirect effect---is a useful addition to any causal mediation claim.

## Parameter Estimates Table

```{r}
#| label: tbl-sr-estimates
#| tbl-cap: "Parameter Estimates for the Structural Regression Model"
#| code-fold: true
```

::: {.callout-note}
## *lavaan* syntax

In *lavaan*:

- `=~` is for measurement (Is measured by)
- `~` is for regression (Is regressed on)
- `~~` is for covariance/variance (Is correlated with)
:::

## Visualization

```{r}
#| label: fig-sr-plot
#| fig-cap: "Structural Regression Model of Self-Oriented Perfectionism, Self-Efficacy, and Academic Procrastination. Standardized estimates are shown."
#| eval: false
```

```{r}
#| label: fig-sr-sempaths
#| fig-cap: "Path diagram from semPaths with semptools styling. Standardized estimates are shown; significant paths are bolded."
#| fig-width: 9
#| fig-height: 5.6
#| code-fold: true
```

## Reporting

### Sample Write-up

. . .

## References

Chen, G., Gully, S. M., & Eden, D. (2001). Validation of a new general self-efficacy scale. *Organizational Research Methods*, *4*(1), 62–83. <https://doi.org/10.1177/109442810141004>

Kline, R. B. (2023). *Principles and practice of structural equation modeling* (5th ed.). Guilford Press.

MacKinnon, D. P., Lockwood, C. M., & Williams, J. (2004). Confidence limits for the indirect effect: Distribution of the product and resampling methods. *Multivariate Behavioral Research*, *39*(1), 99–128. <https://doi.org/10.1207/s15327906mbr3901_4>

Seo, E. H. (2008). Self-efficacy as a mediator in the relationship between self-oriented perfectionism and academic procrastination. *Social Behavior and Personality*, *36*(6), 753–764. <https://doi.org/10.2224/sbp.2008.36.6.753>

Williams, M. N., & Edwards, S. R. (2021). Conceptual replication of Seo (2008): Self-efficacy as a mediator in the relationship between self-oriented perfectionism and academic procrastination. *Current Issues in Personality Psychology*, *10*(2), 98–111. <https://doi.org/10.5114/cipp.2021.110948>
